\hamstr (\emph{H}idden M\emph{a}rkov \emph{M}odel based \emph{S}earch for
or\emph{t}hologs using \emph{R}eciprocity) implements a graph-based approach
using hidden Markov models (HMMs, explained in \autoref{sec:hmms}). It is aimed
specifically at searching for orthologs in \emph{expressed sequence tag} (EST)
data\footnote{National Center for Biotechnology Information: EST fact sheet
(\url{http://www.ncbi.nlm.nih.gov/About/primer/est.html}, revised: March 29,
2004)}, which are sequenced using copy DNA (cDNA) libraries in a process called
\emph{shotgun sequencing}. The cDNA is synthetisized from mRNA and therefore
contains no introns. Like a transcriptome, EST data are incomplete, i.e.,
contain only nucleotide sequences of the genes that were expressed at the time
of RNA preservation. Due to the way they are sequenced and assembled, EST data
can be redundant and fragmented as well, which is why methods for orthology
prediction in genomic data cannot be applied, and approaches like described
above have to be used.

\hamstr cannot be used to perform \emph{de novo} orthology assessment. Instead,
it relies on pre-computed ortholog groups that data from the query EST library
is clustered to. These so-called \emph{core orthologs} are compiled from
existing databases such as OrthoDB or Inparanoid and must be of species for
which a completely sequenced and annotated genome and proteome is available.
From each \emph{ortholog group} (OG, a set of genes that are known to be
one-to-one orthologous), a HMM is generated: each HMM is a probabilistic
representation of the amino acid sequence of a gene. 

\hamstr uses a three-step triangulation strategy as depicted in
\autoref{fig:hamstr-triangulation}: the EST sequence space is searched with the
HMMs for matches (\autoref{fig:hamstr-triangulation}a). To verify orthology in
the HMM hits, each hit sequence is compared with BLASTP \citep{altschul1997} to
the proteome of a so-called reference taxon
(\autoref{fig:hamstr-triangulation}b). Note that the proteome, itself being
completely sequenced and annotated, contains the amino acid sequences that are
known to be orthologous. If the BLAST search hits a sequence that was used to
generate the HMM (\autoref{fig:hamstr-triangulation}c), then orthology is
assumed and the contig in question is added to the ortholog cluster. Otherwise,
it is discarded.

\todo{more elaboration!}

\input{inc/intro/fig-hamstr-triangulation}

% enable this if you need it, it makes compilation a bit slower
%\input{inc/intro/hamstr-flowchart}

\hamstr is written in the Perl programming language for a Linux or UNIX
operating system. It makes use of tools that are part of those environments, and
is therefore not portable to other platforms.
