\label{sec:hmmtest}
To avoid statistical bias by uneven phylogenetic representation when creating a
hidden Markov model (HMM), the HMMER3 package \citep{eddy2009} uses a
position-based weighting scheme as applied by \citet{henikoff1994} by default:
The sequence distances are not calculated based on the sequences as a whole, but
use a diversity measure for each position in the alignment. From the paper:

\begin{quote}
	A simple method to represent the diversity at a position is to award each
	different residue an equal share of the weight, and then to divide that
	weight equally among the sequences sharing the same residue. So, if in a
	position of a multiple alignment, $r$ different residues are represented, a
	residue represented in only one sequence contributes a score of $l/r$ to that
	sequence, whereas a residue represented in $s$ sequences contributes a score
	$l/rs$ to each of the $s$ sequences. For each sequence, the contributions
	from each position are summed to give a sequence weight.
\end{quote}

Thus, if two sequences are very similar in a particular domain, the Henikoff
weighting scheme would penalize that by weighting the sequences down
position-wise. Highly diverse positions, on the other hand, would receive a
bonus. 

In comparison to two species $a$ and $b$ from the same family, the remotely
related species $a$ and $c$ have more divergent sequences. Highly conserved
domains may still be similar, but for the most part, more divergence is
expected. Under the Henikoff weighting scheme, closely related domains will
receive a penalty, while divergent ones are upweighted. Because $a$ and $b$ are
expected to have more positions in common, which results in downweighting of a
larger percentage of their entire sequences, these two sequences are each
``worth'' less than the more remotely related sequence $c$. Two identical
sequences would each receive half the weight that one sequence would.
