\subsection{Algorithm outline}

\subsubsection{Analysis}

The analysis algorithm is as follows:

\begin{enumerate}
	\item Get configuration, initialize global variables.
	\item Check the following:
	\begin{enumerate}
		\item Do the user settings make sense? If not, exit.
		\item Are the paths to the input file and the programs correct? If not, exit.
		\item Do the output directories exist? If not, create them.
		\item Does the database structure exist? If not, exit.
	\end{enumerate}
	\item Backup old results, if requested.
	\item Clear old results from file system and database, if requested.
	\item Create the HMMs, if they do not exist.
	\item Create the BLAST database of all core taxa, if it does not exist.
	\item Configure the hmmsearch and BLAST modules.
	\item Translate the input file in all six reading frames, if it is nucleic acid data.
	\item Load the translated sequences into the database, generating unique SHA256
		digests on the way.
	\item Get all translated sequences of the target species from the database.
		Sequence identifiers are now SHA256 digests. Write the sequences into a
		Fasta file.
	\item For each HMM, do the following:
	\begin{enumerate}
		\item Search the translated sequences with this HMM. Read the tabular search
			report and store it in the database. Skip to the next HMM if no hits were
			obtained.
		\item For each hit, do the following:
		\begin{enumerate}
			\item Get the hit sequence from the database and write the relevant
				subsequence to a Fasta file. This information has been gained from the
				hmmsearch report file.
			\item Search this sequence against the BLAST database. Read the tabular
			search report and store it in the database. Skip to the next HMM if no
			hits were obtained.
		\end{enumerate}
	\end{enumerate}
	\item Print some statistics and exit.
\end{enumerate}

\subsubsection{Reporting}

The reporting algorithm is as follows:

\begin{enumerate}
	\item Get configuration, initialize global variables.
	\item Get a list of ortholog group IDs and their associated ortholog sequences
		from the database.
	\item Get all results in the form:
	\begin{lstlisting}
	hmmsearch_evalue => {
		ortholog_group_id => [
			reciprocal_hit,
			reciprocal_hit,
		],
		ortholog_gene_id => [
			reciprocal_hit,
			reciprocal_hit,
		],
	}
	\end{lstlisting}
	\item Sort the e-values in ascending order.
	\item Starting with the lowest hmmsearch e-value, do the following:
	\begin{enumerate}
		\item For each ortholog group that has a hit with this e-value, do the following:
		\begin{enumerate}
			\item Check whether this ortholog group is present in the list that was
				generated in step 2. If not, skip to the next group (this ortholog group
				has already been assigned a transcript).
			\item Sort the reciprocal hits by BLAST e-value in ascending order.
			\item For each reciprocal hit, do the following:
			\begin{enumerate}
				\item If the reciprocal search hit a sequence that is in this ortholog
					group, and this transcript has not been assigned previously, then this
					is a valid match. Otherwise, skip to the next reciprocal hit unless
					the ``soft threshold'' has been reached (in that case, skip to the
					next ortholog group).
				\item Assign this transcript to this ortholog group. Neither can be
					assigned again.
			\end{enumerate}
		\end{enumerate}
	\end{enumerate}
\end{enumerate}

