During testing, MySQL performance became a bottleneck because of the re-indexing
issues of MyISAM, which take increasingly long with growing table size
(\autoref{sec:mysql-performance}. A solution to this issue is to partition the
tables ``ests'', ``hmmsearch'', and ``blast'' into individual tables for each
query dataset. Transcriptome sequence data as generated within the 1KITE project
have a number of sequences in the order of $5\times10^5$, which amounts to the
same number of records in the database. That means the database in its current
form and without administrative access to server variables can only store up to
about ten query datasets before it becomes slow.

Not only the indexing strategy and the query design, but also the database
structure plays a major part in terms of performance. Especially the tables
``hmmsearch'' and ``blast'' become excessively large after a few analyses: with
an ortholog set of 4,000 OGs, if each of the 4,000 HMM searches obtains an
average of 20 hits during the HMM search, and each of the 20 BLAST searches
obtain an average of 50 hits, this amounts to $4,000 \times 20 = 8 \times 10^4$
rows in the table ``hmmsearch'' and $4 \times 10^6$ rows in the table ``blast''.
The rows themselves do not contain much data (only 131 B on average), but due to
the large number of records, InnoDB performance with the current schema does not
scale well. Insertion of new data does not become noticably slow; however,
deleting old analysis results takes very long. The InnoDB cluster index
physically orders the table based on the primary key or the first unique key it
can utilize. When one row is removed, the entire table is reordered on disk for
speed and defragmentation.  With increasing table size, this operation takes
exponentially long (see \todo{data!}).

The simplest and most performant solution to this problem is to create
individual tables for each query taxon. Dropping or truncating (emptying) a
table and recreating it is much faster than deleting a large portion out of a
huge table, especially when there are indices involved. Implementing this
partitioning is planned for the next version of \pname.
